Nature or Nurture? Collisionless Evolution of Galactic Disc-Halo Systems

### 2019-05-09 Thesis/Background.tex
Writing for May 9

\chapter{A Dynamical Recipe for Cosmological Disks}\label{ch:Background}

\section{Physical Motivation of Modeling}

\section{Time Evolution of Collisionless Systems}
\subsection{The Collisionless Boltzmann Equation}
\subsection{Numerical Solutions}

\section{Phase Space, Equilibrium, and Initial Conditions}
\subsection{Jeans Modeling}
\subsection{DF-based Models and the Strong Jeans Theorem}
\subsection{Action-Angle Variables}
\subsection{DFs as Functions of Actions}

\section{Simulation Analysis: Paradigms and Tools}
\subsection{WKB Wave Analysis}
\subsection{Time Series Filtering}
\subsection{MCMC}

\section{Cosmology and Implications for Galaxy Studies}
\subsection{Basic FRW Cosmology}
\subsection{Extension of Numerical Methods}
\subsection{Sampling Initial Conditions for Cosmological Simulations}
\subsection{Cosmological Substructure and Halo Triaxiality}
\subsection{Identifying Substructure in Simulations}

>>>>

\chapter{A Dynamical Recipe for Cosmological Disks}\label{ch:background}

This chapter provides an overview of the background theory needed to understand subsequent chapters. In Chapter~\ref{ch:introduction}, we had a lot of discussion about simulations of galaxies being critical to having a theoretical understanding of cosmology and galaxy formation. The word ``simulation" was used without fully contextualizing its significance and meaning. We provide that context in \S\ref{sec:motivation}, where we describe what is actually being done when a simulation is run. In \S\ref{sec:galaxy_ics}, we talk about how to apply this theory to study the evolution of equilibrium galaxies. We also spoke at length about $\Lambda$CDM cosmology and how it is critical to explain the evolution of galaxies on a global scale. We talk about cosmology in detail in \S\ref{sec:cosmology}, and explain how this specifies an extended view of the discussion in \S\ref{sec:motivation}. Lastly, we talk about various techniques relied upon in the analysis of simulation in \S\ref{sec:analysis_of_sims}. This includes how we identify cosmic substructure and the techniques we use for analyzing the time-evolution of disk-derived quantities. This chapter summarizes the very basics of the models used in this thesis. It is by no means a comprehensive account of galactic dynamics. In the course of discussion, we will point the reader to more detailed accounts of the topics being discussed.

\section{Physical Motivation of Modeling} \label{sec:motivation}

The goal of this section is to convey to the reader what we mean by a simulation of a galactic system.

\subsection{Characterizing of Self-Gravitating Systems}

The baryonic mass of the Milky Way is largely concentrated in its stars. To first order, the dynamical behavior of the Milky Way is determined by stellar material and dark matter \citep{BM}. While the evolution of gas is governed by the Euler equations with an equation of state, modelling in galactic dynamics requires that we understand how stars and dark matter behave.

Unlike gas, which can exhange thermodynamic energy with itself, stars and dark matter interact only through gravity. Unlike the electromagnetic force, gravity is a long range force. In fact, the majority of the contribution to the forces on stars in galaxies comes from far outside their immediate neighborhoods \citep{BT}. This has a number of interesting implications. In classical thermodynamics, we can derive the macroscopic

\subsection{The Collisionless Boltzmann Equation}
\subsection{Numerical Solutions}

\section{Phase Space, Equilibrium, and Initial Conditions} \label{sec:galaxy_ics}
\subsection{Jeans Modeling}
\subsection{DF-based Models and the Strong Jeans Theorem}
\subsection{Action-Angle Variables}
\subsection{DFs as Functions of Actions}

\section{Cosmology and Implications for Galaxy Studies} \label{sec:cosmology}
\subsection{Basic FRW Cosmology}
\subsection{Extension of Numerical Methods}
\subsection{Sampling Initial Conditions for Cosmological Simulations}

\section{Simulation Analysis: Paradigms and Tools} \label{sec:analysis_of_sims}
\subsection{Cosmological Substructure and Halo Triaxiality}
\subsection{Identifying Substructure in Simulations}
\subsection{WKB Wave Analysis}
\subsection{Time Series Filtering}
\subsection{MCMC}
\bibliographystyle{apalike}
\bibliography{bibliography_background}

### 2019-05-09 Thesis/Introduction.tex
Writing for May 9

\chapter{A Foray into Galactic Dynamics}
\newpage
%\section{Section}
%\subsection{SubSection}
%\subsubsection{SubSubSection}
%\paragraph{Paragraph}
%\subparagraph{SubParagraph}

@@

The aforementioned simulations, although more realistic than single-satellite encounters, have three main problems.  First, they don't allow for halo triaxiality.  Secondly, the disc is initialized at its present-day mass whereas real stellar discs form over several Gyr.  Finally, the subhalo populations are inserted into the halo by hand.  While the mass and spatial distributions of the subhalos are motivated by cosmological simulations, they may not capture important properties of realistic halos. This has historically been solved by modifying a traditional \textit{ab initio} simulation. The zoom-in technique (a Monte Carlo adaptive mesh) is used to study the evolution of a disc in a single halo at high resolution.  This is accomplished by resampling the initial conditions at a sequence of higher resolutions after identifying an area of interest. With sufficiently realistic feedback, impressively realistic results can be obtained. For instance, \citet{gomez_2017} presented a study of vertical structure in such simulations. While highly realistic, the \textit{ab initio} zoom-in approach still suffers from the inability to fine-tune galaxy dynamics in the same halo to perform a truly systematic study. In an effort to address these particular shortcomings, several groups have attempted to grow a stellar disc in a cosmological halo.

@@

@@

We proceed by introducing conformance checking and discussing related work in the next
chapter. We discuss the Alloy language and the Alloy Analyzer tool in
Chapter~\ref{ch:Alloy}.  Chapter~\ref{ch:Embee1} describes our Embee tool, from the
user's perspective, with several running examples.  Implementation details and the
analysis of the tool are presented in Chapter~\ref{ch:Embee2}.
Chapter~\ref{ch:Conclusion} concludes and outlines future work.

>>>>

\chapter{A Foray into Galactic Dynamics} \label{ch:introduction}
\newpage
%\section{Section}
%\subsection{SubSection}
%\subsubsection{SubSubSection}
%\paragraph{Paragraph}
%\subparagraph{SubParagraph}

@@

The aforementioned simulations, although more realistic than single-satellite encounters, have three main problems.  First, they don't allow for halo triaxiality.  Secondly, the disc is initialized at its present-day mass whereas real stellar discs form over several Gyr.  Finally, the subhalo populations are inserted into the halo by hand.  While the mass and spatial distributions of the subhalos are motivated by cosmological simulations, they may not capture important properties of realistic halos. This has historically been solved by modifying a traditional \textit{ab initio} simulation. The zoom-in technique (a Monte Carlo adaptive mesh) is used to study the evolution of a disc in a single halo at high resolution.  This is accomplished by resampling the initial conditions at a sequence of higher resolutions after identifying an area of interest. With sufficiently realistic feedback, impressively realistic results can be obtained. For instance, \citet{gomez_2017} presented a study of vertical structure in such simulations. In these simulations, the authors were able to look at several Milky Way-like galaxies in a variety of environments at resolutions that rival isolated galaxy simulations. These realistic studies have opened the door to a wide variety of questions pertaining to the Milky Way's interaction with the cosmological environment.

@@

While highly realistic, the \textit{ab initio} zoom-in approach still suffers from the inability to fine-tune galaxy dynamics in the same halo to perform a truly systematic study. In an effort to address these particular shortcomings, several groups have attempted to grow a stellar disc in a cosmological halo.

% Should add a footnote about redshift

The general idea for the modification is to divide a dark-matter-only (DMO) zoom-in simulation into three phases.  During the first, a cosmological simulation of a region of interest is run, and a halo is identified for study.  A rigid disc potential is then slowly grown in this halo, allowing the halo particles to respond adiabatically to its time-varying potential.  Finally, the rigid disc is replaced by live particles and the final phase of the simulation proceeds with live disc and halo particles.  \citet{debuhr_2012} use this scheme to introduce stellar discs into dark matter halos from the Aquarius Project \citep{springel2008}.  They add a rigid disc potential at redshift $z=1.3$ with a mass parameter for the disc that grows linearly with the scale factor until $z=1$ when it reaches its final value.  The disc potential is initially centered on the potential minimum of the halo and oriented so that its symmetry axis points along either the minor or major axis of the halo.  During the rigid disc phase, the centre of the potential moves due to its interaction with the halo (Newton's 3rd law) so that linear momentum is conserved.  However, the orientation of the disc is held fixed and therefore angular momentum isn't.  To initialize the live disc, \citet{debuhr_2012} approximate the halo potential as a flattened logarithmic potential and then determine the disc distribution function by solving the Jeans equations in axisymmetry.

A number of refinements to this scheme were introduced by \citet{ys_2015}.  In particular, they use the GALIC code to initialize the live disc \citet{YurinSpringelGalic}.  This code aims to find a stationary solution to the collisionless Boltzmann equation by adjusting the initial particle velocities so as to minimize a certain merit function.  In \citet{ys_2015} the initial disc distribution function was assumed to depend on two integrals of motion, the energy $E$ and angular momentum $L_z$.  Consistency then required that they use an axisymmetrized approximation to the halo potential.

\citet{bauer2018a} improved on the scheme even further by increasing the realism of the growth phase by modeling the disc as a rigid body. The results they found were qualitatively similar to \citet{debuhr_2012} and \citet{ys_2015}  in terms of disc dynamics. The live phase was also improved by using the \textsc{GalactICS} code which actually samples a three-integral DF. It is using this tool that this thesis is constructed upon.

@@

The thesis is organized in the following manner. In Chapter~\ref{ch:background}, we reduce the necessary theoretical background to understand this work to a succinct summary. Following this, in Chapter~\ref{ch:paper_i}, we explore the first peer-reviewed paper on which this thesis is based. It outlines the disk insertion scheme and comments on how the dynamic nature of the Milky Way's stellar disk should affect its inner halo. Chapter~\ref{ch:paper_ii} presents the second published paper which focuses on the question of how stable disks are against forming bars in a cosmological setting. The results discussed in Chapter~\ref{ch:paper_ii} are contextualized with the results of isolated galaxy simulations.
